---
title: "Title it “Working with Spatial Data in R"
format: html
---
```{r}
#| message: false
library(readr)
library(here)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
```


# Load File from KNB
```{r}
knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"

#download the zip file
download.file(url = knb_url, destfile = '../data/shapefile_demo_data.zip')

#unzipping
unzip('../data/shapefile_demo_data.zip', exdir = '../data')

#remove the zip file
file.remove('../data/shapefile_demo_data.zip')
```

## Exploring the data using plot() and st_crs()

```{r}
# read in shapefile using read_sf()
ak_rgns_sf <- read_sf('../data/ak_regions_simp.shp')

# quick plot
plot(ak_rgns_sf)
```
```{r}
#Class of an object
class(ak_rgns_sf)

head(ak_rgns_sf)

glimpse(ak_rgns_sf)
```
# Coordinate Reference System (CRS)

```{r}
# what is the coordinate system of a project
st_crs(ak_rgns_sf)
```
## Transform your geospatial data from one coordinate system to another
```{r}
ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

st_crs(ak_rgns_3338_sf)

plot(ak_rgns_3338_sf)
```
## sf and dplyr

```{r}
# return the name of all the columns
colnames(ak_rgns_3338_sf)

ak_rgns_3338_sf %>% 
  select(region)
```
```{r}
# to find out the names of the different regions in the dataset
unique(ak_rgns_3338_sf$region)

ak_rgns_3338_sf %>% 
  filter(region == "Southeast")
```
## Spatial joins
```{r}

#Read in the population data from a csv and turn it into an sf object
pop_df <- read_csv('../data/alaska_population.csv')

pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = F) #not removing the lat and lng columns

head(pop_4326_sf)
class(pop_4326_sf)
```

#Use a spatial join (st_join()) to assign each city to a region

```{r}
# we need to change the coordinates to match between the two data.frames
pop_3338_sf <- st_transform(pop_4326_sf, 
                            crs = 3338)

#then we can join them together
pop_joined_sf <- st_join(pop_3338_sf, 
                         ak_rgns_3338_sf, 
                         join = st_within) 

head(pop_joined_sf)
```

#Use group_by() and summarize() to calculate the total population by region
```{r}
pop_region <- pop_joined_sf %>% 
  as.data.frame() %>% #will remove geometry, sf --dataframe
  group_by(region) %>% 
  summarize(total_pop = sum(population))

head(pop_region)
class(pop_region)
```

## population by region
```{r}
pop_region_3338  <- left_join(ak_rgns_3338_sf,
                              pop_region,
                              by = "region")
# plot of the population by region
plot(pop_region_3338["total_pop"])


```

## population by managment area
```{r}
pop_mgmt_3338_sf <- pop_region_3338%>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop), do_union = F)

#If we don’t want to combine geometries, we can specify do_union = F as an argument.

plot(pop_mgmt_3338_sf["total_pop"])
```
## Save the spatial object you created using write_sf()
```{r}
write_sf(pop_region_3338,"../data/ak_regions_populations.shp")
```

# Visualize with ggplot

```{r}
ggplot(pop_region_3338) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) + #this allows to spcify the type of value format (in this case, instead of exponential)
    theme_bw()
```
## add rivers in Alaska and plot them by size
```{r}
rivers_3338_sf <- read_sf("../data/ak_rivers_simp.shp")
st_crs(rivers_3338_sf)

ggplot()+
  geom_sf(data = pop_region_3338,
          aes(fill = total_pop))+
  geom_sf(data = pop_3338_sf,
          size = 0.5)+
  geom_sf(data = rivers_3338_sf,
          aes(linewidth = StrOrder))+ #Strahler stream order (larger number = larger river)
  scale_linewidth(range = c(0.05, 0.5), guide = "none") +
  labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()


```

# Incorporate base maps into static maps using ggspatial
Pseudo-mercator (EPGS 3857)

```{r}
#changing the EPGS
pop_3857_sf <- st_transform(pop_3338_sf, 
                         crs = 3857)

# plotting
ggplot(data = pop_3857_sf) +
    ggspatial::annotation_map_tile(type = "osm", #"osm" are the tiles
                                   zoom = 4, #type of zoom
                                   progress = 'none') + # higher zoom values are more detailed
    geom_sf(aes(color = population),
            fill = NA) +
    scale_color_continuous(low = "darkkhaki",
                           high =  "firebrick",
                           labels = comma)
```

# Visualize sf objects with leaflet
```{r}
# first need to save our custom projection for leaflet

epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS", # what projection library to use
    code = "EPSG:3338", # CRS code
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",# this is from the definition of the proj.
    resolutions = 2 ^ (16:7) #dividing the scale in 8*2
)

#You might notice that this looks familiar! The syntax is a bit different, but most of this information is also contained within the crs of our shapefile:
st_crs(pop_region_3338)
```

## Since leaflet requires that we use an unprojected coordinate system, let’s use st_transform() yet again to get back to WGS84.

```{r}
pop_rgn_4326_sf <- pop_region_3338 %>% 
    st_transform(crs = 4326)
st_crs(pop_rgn_4326_sf)

pop_4326_joined <- pop_joined_sf %>% 
    st_transform(crs = 4326)
st_crs(pop_4326_joined)
```

## make leaflet map

```{r}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_rgn_4326_sf,
                fillColor = "gray",
                weight = 1)

m
```

## make a better map
```{r}
pal <- colorNumeric(palette = "Reds", 
                    domain = pop_rgn_4326_sf$total_pop)
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop), #here we access the palette that we just created (using ~)
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~ region
    ) %>% 
  addCircleMarkers(
        data = pop_4326_joined,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326_joined$city, ", population ", comma(pop_4326_joined$population))
    ) %>% 
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_4326_sf$total_pop),#creating a range using the range function, on a det. column
        title = "Total Population"
    )

m
```

