---
title: "Tidying and Wrangling"
format: html
---

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(readr)
```

# 1. Data sources

## 1.1 Read in our data

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## 1.2 Explore

```{r}
#| message: false
colnames(catch_original)
head(catch_original)
summary(catch_original)
unique(catch_original$Region)
```

# 2. Cleaning & reshaping

## 2.1 Using the select function

```{r}
#Select () to keep the coluns that want
catch_data <- catch_original %>% 
  select( Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)

catch_data_2 <- catch_original %>% 
  select( -All, -notesRegCode)
head(catch_data)

```

## 2.2 Check columns type

```{r}
glimpse(catch_data)
```

## 2.3 Changing column contents with mutate ( )

```{r}
#Change Chinook to numeric
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))
```

```{r}
is.na(catch_clean$Chinook)

#get the location of the NA value
i  <-which (is.na(catch_clean$Chinook))
i

# look at the row in the data
catch_data[i,]

# use mutate to fix
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                           true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook))
head(catch_clean)

# To check for the row (401) that had the issue:
catch_clean[i,]
```

## 2.4 Changing file shape using pivot_longer and pivot_wider ( )

```{r}
catch_long <- catch_clean %>%
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch")
head(catch_long)

# Using the wide function., it comes back to the previous dataset:

catch_wide <- catch_long %>%
  pivot_wider(names_from = "species",
              values_from = "catch")
head(catch_wide)

```

## 2.5 Rename the columns with rename ( )

```{r}
catch_long <- catch_long %>%
  rename(catch_thousands = catch)

head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)

head(catch_long)
  
```

#removing the "catch_thousands" column

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)

head(catch_long)
```

# 3. Summary statistics

```{r}
#. get the average by region
mean_region <- catch_long %>%
    group_by(Region) %>% #we pipe again because it is for each of those groups
    summarize(mean_catch = mean(catch))

head(mean_region)

# get the number of rows for each group
n_region <- catch_long %>%
    group_by(Region) %>%
    summarize(n = n()) #function n will get the counts

head(n_region)

```

## 3.1 Filter rows using filter( )

```{r}
sse_catch <- catch_long %>%
    filter(Region == "SSE")

head(sse_catch)

## try filtering where catch column is 1 million fish

catch_million_fish <- catch_long %>%
    filter(catch > 1000000)

head(million_fish)

## Chinook from SSE data (two at a time)
chinook_sse <- catch_long %>%
    filter(Region == "SSE",
           species == "Chinook")
```

## 3.2 Sorting data with arrange ( )

```{r}
#using desc()
mean_region <- catch_long %>%
    group_by(Region) %>% 
             summarise(mean_catch = mean(catch)) %>% 
               arrange(desc(mean_catch))
head(mean_region)
```

## 3.3 Combining datasets together with join ( )

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)

catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
                          
head(catch_joined)
```

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
  rename(Region = code, Region_name = mgmtArea)

catch_joined <- left_join(catch_joined, region_defs, by = c("Region"))
head(catch_joined)

```

## 3.4 Split and combine values in columnns

```{r}
#Creating a data fram with three codes and values
sites_df  <- data.frame(site = c("HAW-101",
                                 "HAW-103",
                                 "OAH-320"))
head(sites_df)

split_df  <- sites_df %>% 
  separate(site, c("island", "site_number"), sep = "-")
```

```{r}
dates_df  <- data.frame(
  year = c("1930", "1930", "1932"),
  month = c("12", "12", "12"),
  day = c("14", "15", "16")
)

dates_unite  <- dates_df %>%
  unite(col = date, year, month, day, sep = "/")
head(dates_unite)
```

```{r}
mean_region  <- catch_original %>% 
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                           true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year),
               names_to ="species",
               values_to = "catch") %>% 
  mutate(catch = catch * 1000) %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
```

# 4. Save our results with write_csv()

```{r}
library(here)
write_csv(mean_region, here("data/mean_catch_by_region.csv"))
```
